---
title: "HW 5  - Network visualization"
author: "Narek Sahakyan"
date: "11/20/2019"
output: pdf_document
---
Dear Mr. Madoyan, I have decided to 
analyze the transfer market as a network.
The data contains 4700 transfers from season 2000-2001
upto 2018-2019. As the number of teams is huge, I have decided
to filter out the deals between the teams from top 10 leagues.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages_list <- c("igraph","network","circlize","dplyr","GGally")

install_or_call <- function(list = packages_list){
  installed <- installed.packages()[,"Package"]
  for( package in packages_list ){
    if(!(package %in% installed)){
      install.packages(package)
    }
    do.call(library, list(package))
  }
}
install_or_call()
transfers <- read.csv("transfers.csv",stringsAsFactors = F)
```


```{r, echo=F}
top10 <- c("LaLiga", "Serie A", "1.Bundesliga", "Ligue 1", "Premier League",
          "Liga NOS", "Premier Liga", "MLS", "Eredivisie", "Série A")


by_lg <- transfers %>%
  filter(League_from %in% top10 & League_to %in% top10) %>%
  group_by(League_from, League_to) %>%
  summarise(Count = n(), Mean_Fee = round(mean(Transfer_fee) / 1000000),
            Mean_Age = round(mean(Age))) %>%
  filter(Count >= 10) %>%
  arrange(desc(Mean_Fee / Count))

lg_igraph <- igraph::graph_from_edgelist(as.matrix(by_lg[,1:2]), directed = T)
lg_igraph <- set_edge_attr(lg_igraph, "count", value = by_lg$Count)
lg_igraph <- set_edge_attr(lg_igraph, "mean.price", value = by_lg$Mean_Fee)
lg_igraph <- set_edge_attr(lg_igraph, "mean.age" , value = by_lg$Mean_Age)

lg_net <- intergraph::asNetwork(lg_igraph)

count <- lg_net %e% "count"
count <- (count - min(count))/ (max(count) - min(count))

# Visualizing with network
# set.seed(1)
# plot(lg_net, label = network.vertex.names(lg_net),edge.lwd = count * 30,
#      mode = "circle", displaylabels = T, edge.arrow.mode = 3)

l <- layout_in_circle(lg_igraph)

count1 <- E(lg_igraph)$count
count1 <- (count1 - min(count1))/ (max(count1) - min(count1))

plot(lg_igraph, layout = l, edge.arrow.mode = 3, edge.width = count1 * 20,
     edge.curved = T)

```
As we can see the top 5 leagues are stronlgy interconnected to each other,\
as most of the transfers take place between the teams from the top 5 \
We can also see loops, meaning that many transfers take place in the same league \
It is visible that Premier League teams are the most attracting as the majority of the trasnfers \
happen end up having a buyer club from the Premier League. The league also has high positions \
in terms of inside the league transfers, alongside with Seria A. Seria A and Premier League have higher \
number of inside league transfers in comparison with the other represenatives of the top 5(France, Spain, Germany)


```{r, echo = F}
as_m <- as.matrix(lg_net, matrix.type = "adjacency")

chordDiagram(as_m)
````

Almost the same picture is expressed in this visualization, \b
however we can clearly notice that the teams from the non top 5 leagues \
such as Portugal(Liga Nos, Eredivise) mostly sell their players to top 5 leagues teams\
rather than buying them. Most of the incoming transfers are from non popular leagues such \b
as the Brazilian league. I know that there are two teams in Portugues league that are experts \
in buying young Brazilian players and then selling them to top leagues. On the other side \
Ajax, one of the most famous teams of in the world in terms of its own academy. \
Following Ajax's principle most of the Eredivisie teams rise and sell their best players \
However, we can see that the Russian League has relatively higher number of incoming transfers\b
in comparison with the noted countries. In my opinion the high salaries in Russian teams attract \
the players, or most of them are about to finish their career. As with the age, the abilities of the \
players seem to get worse, they often switch to weaker teams to gain more gaming practice or just make money\