---
title: "Homework 3"
author: "Narek Sahakyan"
date: "27 October, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SportsAnalytics270)
packages_list <- c("ggplot2", "dplyr")

install_or_call <- function(list = packages_list){
  installed <- installed.packages()[,"Package"]
  for( package in packages_list ){
    if(!(package %in% installed)){
      install.packages(package)
    }
    do.call(library, list(package))
  }
}

install_or_call()
data("f_data_sm")
data("transfers")
transfers <- na.omit(transfers)
```


TOTAL: 50p

Reminder: In football draw is a half win.
          Number of teams per season might change even within given league

Noll-Skully number 

*Your goal is to calculate Noll-Skully number for football
* Chose any 4 leagues from f_data_sm on your own
* Calculate Noll_Skully Number as a whole for all the seasons for the given league. (2p)
```{r}
countries <- c("England","Germany","Spain","Italy")
top_4_data <- f_data_sm %>%
  filter(COUNTRY %in% countries)

final_tables <- function(data,country){
  result <- c()
  for(season in unique(data$SEASON)){
    season_table <- final_table(data, country, season)
    league <- data[data$SEASON == season & 
                     data$COUNTRY == country,]$LEAGUE[1]
    season_table$SEASON = season
    season_table$COUNTRY = country
    season_table$LEAGUE = league
    result <- rbind(result, season_table)
  }
  cols <- colnames(result)
  len <- length(cols)
  result <- result[c("SEASON","COUNTRY","LEAGUE",cols[1 : (len-3)])]
  return(as.data.frame(result))
}

countries_standings <- function(data){
  result <- c()
  for(country in unique(data$COUNTRY)){
    country_standings <- final_tables(data,country)
    result <- rbind(result, country_standings)
  }
  return(result)
}

top_4_standings <- countries_standings(top_4_data)
top_4_standings$TW = round(top_4_standings$W + (top_4_standings$D)/2)
top_4_standings$WR <- top_4_standings$TW / top_4_standings$M
#Win Ratio = number of wins / number of games
#Tunned wins = number of wins + (number of draws)/2
#As already mentioned in the description the number of games played
#each season in football can change even in the same league,
#so in order to have a general metric for summarizing let's take
#the mean of the number of games played in each season as an approximation
#We will use the exact number of games later, 
#when analyzing the CB using season based approaches

#Version 1 teams_count
#Number of teams = number of games / 2 + 1
# teams_count <- top_4_standings %>%
#   group_by(COUNTRY) %>%
#   summarise(MEAN.TC = round(mean( M/2 + 1 )))
# id_s <- 0.5 / sqrt(teams_count)

mean_t <- function(M) {
  return(round( mean( M/2 + 1 ) ))
}

top_4_nsc_v1 <- top_4_standings %>%
  group_by(COUNTRY) %>%
  summarise(NSC = sd(WR) / ( 0.5 / sqrt(mean_t(M)))) %>%
  mutate(V = "V1") %>%
  arrange(desc(NSC))
#Now let's use count three draws as one win
top_4_standings$TW_v2 <- round(top_4_standings$W + (top_4_standings$D)/3)
top_4_standings$WR_v2 <- top_4_standings$TW_v2 / top_4_standings$M
top_4_nsc_v2 <- top_4_standings %>%
  group_by(COUNTRY) %>%
  summarise(NSC = sd(WR_v2) / ( 0.5 / sqrt(mean_t(M)))) %>%
  mutate(V = "V2") %>%
  arrange(desc(NSC))

# As we can see the competitive balance increased for all the leagues
# when using this type of calculations. 
# The increase in the competetive balance
# means that the role of luck gets less decisive. 
# However the changes are very slight and 
# not significant, so I am not sure,
# but I believe that having these facts we can conclude
# that the draws in general do slightly affect the all time CB for these leagues, as 
# when we give them less importance(v1: 2 draws = 1 win, v2: 3 draws = 1 win) 
# the CB is increasing which can be translated into 
# the decrease of the importance of the luck. I find this connection logical
# as if the CB is getting higher when we discard more draws, 
# as we count 3 of them as a win. However when the CB is close to 
# being balanced we expect more draws in general. 
# Now let's imagine two simple examples.
# Suppose team X won 1 game and tied 10 times. 
# If we want to transform the draws into wins using half principle
# We would say that X won 6 of its games, using 1/3 principle 
# we would say that the team won 4 of its games.
# According to our approach, in general the discard of that two "won" games leads to 
# increase in CB(decrease in luck importance). 
# Using 1/3 principle the teams with most wins will get the 
# highest portion of wins, and the teams with more draws 
# will get lower rates in comparison to the rates 
# they would get using half principle(6 vs 4) and 
# this will lead to higher variance in skill(CB). So the teams
# with draws the lower will become this variance in skill,
# which will increase the luck's importance as in 
# general there would be more equal teams and more expected draws.

# I hope I have written something logical :)



#Version 2 matches count
# Let's use the same approach as for version one, 
# but considering the number of games played
# rather than the number of the teams.
mean_m <- function(M){
  return(round(mean(M)))
}

top_4_nsc_v3 <- top_4_standings %>%
  group_by(COUNTRY) %>%
  summarise(NSC = sd(WR_v2) / ( 0.5 / sqrt(mean_m(M)))) %>%
  mutate(V = "V3") %>%
  arrange(desc(NSC))

top_4_nsc <- rbind(top_4_nsc_v1,
                   top_4_nsc_v2,
                   top_4_nsc_v3)
# As we can see all the approaches identified the same ranking 
# of the leagues in terms of CB over all time, so let's check 
# the differences in CB for the leagues for all of the approaches
```

* Plot these numbers together and comment (6p)
```{r}
top_4_nsc %>%
  ggplot(aes(x = reorder(COUNTRY,NSC))) + 
  geom_bar(aes(y = NSC),stat = "identity") + 
  facet_grid(.~V) + 
  labs(x = "COUNTRIES", y = "NOLL SCULLY NUMBER") + 
  ggtitle("CB CALCULATIONS USING DIFFERENT APPROACHES")
# As we can see the differences in the leagues in terms of CB balances remain
# almost the same for V1 and V2, but become more sparse for V3. 
# However this is mainly because of considering 
# the number of matches instead of number of teams.
# I believe that using V3 is more relevant for football.
# As we know in general the 
# Number of Games in a competition = (Number of teams - 1 ) * 2,
# So addition of one team will lead to two more games, two teams = 4 more games and so on. 
# In general, more the games, more the chances of change in CB. 
# Supposing that we calculate ID_S by the formula
# 0.5 / sqrt(FACTOR), where FACTOR 
# is either 1)the number of games or 2)the number of teams.
# Unit increase in 1) leads to double of that increase in 2), 
# the higher the denominator the lower the ID_S 
# which in case will lead to higher CB as CB = SD(WPCT)/ID_S. 
# As number of games are derived 
# from the number of teams(opposite is also true in general) and in general 
# they are more important in CB as CB is more sensitive to it and becomes
# higher when regarding them as factor, 
# In my opinion it is better to use number of games as a FACTOR for ID_S

top_4_nsc %>%
  group_by(V) %>%
  summarise(SD = sd(NSC)) %>%
  arrange(desc(SD))
```

* Now do the same by season (6p)
```{r}
top_4_nsc_v4 <- top_4_standings %>%
  group_by(SEASON, COUNTRY) %>%
  mutate(WR = (W + D/3) / M) %>%
  summarise(NSN = sd(WR) / (0.5 / sqrt(unique(M)))) %>%
  arrange(desc(NSN))
```

* Plot and comment (8P)

```{r}
highest_cbs <- top_4_nsc_v4 %>% 
  group_by(SEASON) %>%
  filter(NSN == max(NSN)) %>%
  arrange(desc(NSN))


narrow_standings <- function(cbs = highest_cbs,standings = top_4_standings){
  result <- c()
  for( index in 1:nrow(cbs) ) {
    competition <- cbs[index,]
    standings <- top_4_standings %>%
      filter(SEASON == competition$SEASON,
             COUNTRY == competition$COUNTRY)
    result <- rbind(result, standings)
  }
  return(result)
}

competitive_standings <- narrow_standings()
# As we take a look at this data frame we can see
# that in these years when the CB was the highest 
# the standings were tense, and the difference in points
# for the clubs is minimal for each country. Even in Italy
# in season 2017/2018 Juve was had only 4 more points than 
# Napoli. In 1999 the most competitive season was in Germany.
# It is interesting that the difference in points was not that
# tense, this is mainly because Germany had the lowest CB over 
# time when not regarding the seasons, meaning that even normal 
# differences in points for Germany seem competitive as most of 
# the time CB is low.

top_4_nsc_v4 %>%
  ggplot(aes(x = SEASON, y = NSN)) + 
  geom_line() + 
  facet_grid(.~COUNTRY) + 
  labs(x = "SEASON", y = "NOLL SCULLY NUMBER") + 
  ggtitle("CB CALCULATIONS OVER TIME IN TOP 4 LEAGUES")
# As we can see currently and over time England seems to have
# the highest competitive balance. We can see that the highest 
# value of CB was in 2019, when Man City won the championship
# by having only one more point than Liverpool. England and Germany
# were only countries in which the season 2018/2019 was more competitive
# than the previous season, whereas In Spain the competitive balance
# rate dropped significantly for season 2018/2019, in Italy the pattern
# was also the same, but it was not that significant and visible as the season
# 2017/2018 was the most competitive.
# In general we can see that the ranking of the leagues in terms of being
# competitive is expressed through the visualization as England is the most competitive
# and Germany has the lowest CB over time in comparison with the others. All of the leagues
# had some peak values(both low and high) and in that period, we can see that the standings
# were accordingly tense or variated.
```

Your goal is to find 2 leading and 2 lagging indicators for those leagues.
Show correlation (on plot and calculating correlation coefficient) between these indicators and Noll-Skully number (or any other competitive balance metric on your choice). 20p

Explain why do you think these variables are leading or lagging. 10p

```{r}

```

